Ricci curvature of the phylogenetic subtree-prune-regraft graph

### 2015-01-21 curvature.tex
statement of degree change lemma

\newattribute{OhOf}{\mathrm{O}}
\newattribute{ThetaOf}{\Theta}
\newattribute{OmegaOf}{\Omega}
\newattribute{ohOf}{\mathrm{o}}
\newattribute{omegaOf}{\omega}
%\newattribute{min}{\text{min}}
%\newattribute{max}{\text{max}}

@@

>>>>

\newattribute{OhOf}{\mathrm{O}}
\newattribute{ThetaOf}{\Theta}
\newattribute{OmegaOf}{\Omega}
\newattribute{ohOf}{\mathrm{o}}
\newattribute{omegaOf}{\omega}
\newattribute{lca}{\text{LCA}}
%\newattribute{min}{\text{min}}
%\newattribute{max}{\text{max}}

@@

\section{Rooted SPR Neighborhoods}

\begin{lemma}
Let $T$ and $S$ be phylogenetic trees such that $\dspr{T,S}=1$, that is $T$ can be transformed into $S$ by moving a subtree $R$ with $k$ leaves.
The corresponding SPR operation moves $R$ from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
If multiple SPR moves transform $T$ into $S$ (i.e. an NNI\todo{lemma}), then assume that $V$ is the aunt of $R$ in $T$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the distance from $U$ to $L$ in $T$, and $b$ the distance from $V$ to $L$ in $T$.
Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degree of $T$ and $S$ differs by:
$$(2k-1)(a-b-1) + 2(i - j).$$

\end{lemma}
\begin{proof}

Proof strategy:

Gain $(2k-1)(a-2)$ moves from the $a$-path to $R$.

Lose $(2k-1)(b-1)$ moves from the $b$-path to $R$.

Gain $2(i-j)$ moves involving the parent of $R$ to $U$ and $V$.

\end{proof}
